
\documentclass[10pt]{article} % For LaTeX2e
%\usepackage{tmlr}

% If accepted, instead use the following line for the camera-ready submission:
%\usepackage[accepted]{tmlr}
% To de-anonymize and remove mentions to TMLR (for example for posting to preprint servers), instead use the following:
\usepackage[preprint]{tmlr}

% Optional math commands from https://github.com/goodfeli/dlbook_notation.
\input{math_commands.tex}

\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{url}
\usepackage{amsmath}


\newtheorem{remark}{Remark}
\newtheorem{definition}{Definition}
\newtheorem{axiom}{Axiom}
\newtheorem{theorem}{Theorem}
\newtheorem{proof}{Proof}

\title{Using the Path of Least Resistance to Explain Deep Networks}

% Authors must not appear in the submitted version. They should be hidden
% as long as the tmlr package is used without the [accepted] or [preprint] options.
% Non-anonymous submissions will be rejected without review.

\author{\name Sina Salek \email salek.sina@gmail.com\\
	\addr Independent Researcher
	\AND
	\name Joseph Enguehard \email joseph.enguehard@robinai.com \\
	\addr Robin AI\\ 10 Devonshire Square, City of London, London EC2M 4YP
}

% The \author macro works with any number of authors. Use \AND 
% to separate the names and addresses of multiple authors.

\newcommand{\fix}{\marginpar{FIX}}
\newcommand{\new}{\marginpar{NEW}}

%\def\month{02}  % Insert correct month for camera-ready version
%\def\year{2025} % Insert correct year for camera-ready version
%\def\openreview{\url{https://openreview.net/forum?id=XXXX}} % Insert correct link to OpenReview for camera-ready version


\begin{document}
	
	
	\maketitle
	
	\begin{abstract}
		Integrated Gradients (IG), a widely used axiomatic path-based attribution method, assigns importance scores to input features by integrating gradients of the model along a straight path from a baseline to the input. While effective in certain cases, we show that choosing straight paths can lead to flawed attributions. In this paper, we identify how these misattributions arise. As a solution, we propose a new approach that considers the input space as a Riemannian manifold and computes attributions by integrating gradients of the model along geodesics. We call our approach \emph{Geodesic Integrated Gradients}. To approximate geodesic paths, we introduce two methods: a \emph{k}-Nearest Neighbour-based approach for smaller models and a Stochastic Variational Inference-based method for larger ones. We formulate an additional axiom to the ones satisfied by IG. We show that this is a desired property for an attribution method to satisfy, and that Geodesic Integrated Gradients is the only method that satisfies it. Furthermore, in our experiments with both synthetic and real world data, we demonstrate that our approach outperforms existing explainability methods including the original IG.\footnote{The code for the attribution methods and reproducing the experiments is provided on \href{https://github.com/sina-salek/geodesic-ig}{https://github.com/sina-salek/geodesic-ig}}
	\end{abstract}
	\input{introduction}
	\input{method}
	\input{experiments}
	\input{related_work}
	\input{discussion}
	
	\bibliography{bib}
	\bibliographystyle{plainnat}
	
	\newpage
	\appendix
	\onecolumn
	\input{appendix}
	
\end{document}